$R^*$: A robust MCMC convergence diagnostic with uncertainty via gradient-boosted machines

### 2020-03-16 write_up/ms.tex
added algorithm blocks

@@

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$ where $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \cite{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm X shows how this statistic is constructed.

@@

A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$. For each sample, $s$, in our testing set, our GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a single chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^s=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this sampling process for each sample in the testing set, generating $\boldsymbol{a}=(a^1,a^2,...,a^{N_\text{test}})$, whose average yields a single $R^{*\{i\}}=\bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$. Algorithm Y shows how to calculate this distribution for a given set of MCMC samples.

>>>>

Add another panel for the joint case

@@

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$ where $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \cite{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives an explicit recipe for calculating $R^*$.

\begin{algorithm}[tb]
	\caption{$R^*$ calculation algorithm}
	\label{alg:R_star}
	\begin{algorithmic}
		\STATE Given $\theta^0$, $\epsilon$, $L$, $\cL, N$:
		\FOR{$m=1$ to $N$}
		\STATE Sample $k^0\sim\N(0, M)$.
		\STATE Set $\theta^m\leftarrow \theta^{m-1}, \tilde\theta\leftarrow \theta^{m-1}, \tilde k\leftarrow k^0$.
		\FOR{$i=1$ to $L$}
		\STATE Set $\tilde\theta, \tilde k \leftarrow \LF(\tilde\theta, \tilde k, \epsilon)$.
		\ENDFOR
		\STATE With probability $\alpha = \min\{1, \exp(H(\theta^{m-1}, k^0)-H(\theta^m,\tilde k))\}$
		set $\theta^m\leftarrow\tilde \theta$, $k^m\leftarrow -\tilde k$.
		\ENDFOR
		\STATE
		\STATE {\bf function} $\LF(\theta, k, \epsilon)$
		\STATE Set $\tilde k \leftarrow k +
		(\epsilon/2)\nabla_\theta\cL(\theta)$.
		\STATE Set $\tilde \theta \leftarrow \theta +
		\epsilon M^{-1}k'$.
		\STATE Set $\tilde k \leftarrow \tilde k +
		(\epsilon/2)\nabla_\theta\cL(\tilde\theta)$.
		\RETURN $\tilde \theta, \tilde k$.
	\end{algorithmic}
\end{algorithm}

@@

A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words. For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=\bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

\begin{algorithm}[tb]
	\caption{Procedure to generate a single $R^*$ sample}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given $\theta^0$, $\epsilon$, $L$, $\cL, N$:
		\FOR{$m=1$ to $N$}
		\STATE Sample $k^0\sim\N(0, M)$.
		\STATE Set $\theta^m\leftarrow \theta^{m-1}, \tilde\theta\leftarrow \theta^{m-1}, \tilde k\leftarrow k^0$.
		\FOR{$i=1$ to $L$}
		\STATE Set $\tilde\theta, \tilde k \leftarrow \LF(\tilde\theta, \tilde k, \epsilon)$.
		\ENDFOR
		\STATE With probability $\alpha = \min\{1, \exp(H(\theta^{m-1}, k^0)-H(\theta^m,\tilde k))\}$
		set $\theta^m\leftarrow\tilde \theta$, $k^m\leftarrow -\tilde k$.
		\ENDFOR
		\STATE
		\STATE {\bf function} $\LF(\theta, k, \epsilon)$
		\STATE Set $\tilde k \leftarrow k +
		(\epsilon/2)\nabla_\theta\cL(\theta)$.
		\STATE Set $\tilde \theta \leftarrow \theta +
		\epsilon M^{-1}k'$.
		\STATE Set $\tilde k \leftarrow \tilde k +
		(\epsilon/2)\nabla_\theta\cL(\tilde\theta)$.
		\RETURN $\tilde \theta, \tilde k$.
	\end{algorithmic}
\end{algorithm}
